---
title: "BreedingR: from data to breeding value"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BreedingR: from data to breeding value}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2026)
```

This document walks the full workflow on simulated data, simulated HERE, in the vignette,
so that every number shown has the truth sitting next to it. An example whose result
cannot be checked teaches nothing.

## 1. Simulate a herd

A pedigree with herd structure (few sires, many dams) and genetic values by the
recursion `a = 0.5(a_sire + a_dam) + mendelian`, never by a Cholesky of A, which would
share code with the very thing we are testing.

```{r}
library(BreedingR)

n <- 800
id <- sprintf("a%04d", 1:n)
sire <- dam <- rep("0", n)
for (i in 61:n) {
  sire[i] <- id[sample(1:20, 1)]
  dam[i] <- id[sample(1:(i - 1), 1)]
}
ped <- data.frame(id, sire, dam)

p <- pedigree(ped)   # topological order + Meuwissen-Luo (1992) inbreeding
p
```

```{r}
va <- 0.4; ve <- 0.6           # the truth: h2 = 0.4
a <- numeric(n)
for (i in 1:n) {
  s <- p$sire[i]; d <- p$dam[i]
  mend <- if (!is.na(s) && !is.na(d)) 0.5 - 0.25 * (p$F[s] + p$F[d])
          else if (!is.na(s) || !is.na(d)) 0.75 else 1
  base <- 0
  if (!is.na(s)) base <- base + 0.5 * a[s]
  if (!is.na(d)) base <- base + 0.5 * a[d]
  a[i] <- base + sqrt(mend * va) * rnorm(1)
}
cg <- sample(sprintf("g%02d", 1:10), n, replace = TRUE)
efc <- setNames(rnorm(10), sprintf("g%02d", 1:10))
data <- data.frame(id = p$id, cg = cg,
                    peso = 30 + efc[cg] + a + sqrt(ve) * rnorm(n))
```

## 2. Look before estimating

```{r}
describe(data, "peso", classes = "cg", pedigree = ped)
```

## 3. Fit the animal model

```{r}
fit <- model(peso ~ cg + animal(id), data, pedigree = ped)
fit
```

The truth was `va = 0.4`, `ve = 0.6`. The distance in standard errors is what says
whether the recovery is within expectation; an estimator that hit 0.400 on the nose
would be reading the answer, not estimating it.

`theta` is a named vector, so both halves of the ratio are taken by name: the numerator
is the animal variance and the denominator sums the `var()` entries only, never a
covariance or the `rho` an AR(1) residual adds.

```{r}
v <- fit$theta[grepl("^var[(]", names(fit$theta))]
c(h2_estimated = round(v[["var(animal)"]] / sum(v), 3),
  z_va = round(abs(fit$theta[["var(animal)"]] - va) / fit$se[["var(animal)"]], 2))
```

## 4. Breeding values and accuracy

```{r}
vg <- ebv(fit)
acc <- accuracy(fit, ped)
summary(acc)
plot(vg[order(vg)], acc[order(vg)], pch = ".", cex = 2,
     xlab = "EBV (sorted)", ylab = "accuracy")
```

## 5. Models the formula declares without a dedicated fitter

The layout unit is the covariance group: `group = "g"` puts two random terms into the
same matrix, with the correlation estimated.

```{r, eval = FALSE}
# direct-maternal
model(peso ~ cg + animal(id, group = "g") + maternal(dam, group = "g"), data, ped)

# reaction norm on a Legendre basis
d2 <- cbind(data, legendre(data$idade, order = 1, limits = c(150, 190)))
model(peso ~ cg + rn(id, base = c("phi0", "phi1")) + pe(id), d2, ped)

# indirect genetic effect (the associative model of Muir & Schinckel, 2002)
model(peso ~ cg + animal(id, group = "g") + indirect(id, pen = "baia", group = "g"),
        data, ped)

# single step, with the 0/1/2 genotype matrix
model(peso ~ cg + animal(id), data, ped, genotypes = list(ids = gids, m = M))
```

## 6. The Bayesian half

The same model, sampled instead of maximized. The posterior means should land near the
REML estimates — and they carry the whole distribution with them.

```{r}
gb <- gibbs(peso ~ cg + animal(id), data, pedigree = ped,
            n_iter = 3000, burnin = 500, thin = 5)
data.frame(component = colnames(gb$samples),
           posterior = round(unname(gb$mean), 3),
           reml = round(unname(fit$theta), 3))
```

## 7. Genomics without ever building G

For genomic examples the package ships its own generator, `simulate_breeding()`, which
gene-drops haplotypes so the genotypes are CONSISTENT with the pedigree it emits. The
single step runs three ways: exact, `apy_core=` (a global core), or `vecchia_k=` (each
animal conditions on its own k strongest previous relationships — Henderson's (1976)
A^-1 is this recursion with the parents as the set). And `snp_blup()` drops G entirely: markers
enter as equations and the system is solved by conjugate gradients at given components.

```{r}
s <- simulate_breeding(n_founders = 40, n_generations = 2,
                       offspring_per_generation = 80, h2 = 0.4,
                       n_markers = 300, seed = 7)
fg <- model(y ~ cg + animal(id), s$data, s$pedigree, genotypes = s$genotypes)
fs <- snp_blup(y ~ cg + animal(id), s$data, s$pedigree, s$genotypes,
               theta = unname(fg$theta))
cor(ebv(fg, "animal"), ebv(fs, "animal")[names(ebv(fg, "animal"))])
```

The two paths differ by a declared detail (the affine adjustment of G* that `snp_blup()`
does not apply), so the correlation is near 1 without being 1. PLINK files enter through
`read_plink()`, and `qc_genotypes()` filters call rate, MAF and HWE while REPORTING what
each filter removed.

## 8. The selection side

```{r}
tbv <- s$tbv[names(ebv(fg, "animal"))]
idx <- selection_index(list(y = ebv(fg, "animal")), c(y = 1))
dr <- rank_drift(ebv(fs, "animal"), ebv(fg, "animal"), top = 50)
c(cor_index_tbv = round(cor(idx[names(tbv)], tbv), 2),
  drift_spearman = round(dr$spearman, 4), retention_50 = dr$retention)
```

`mc_study()` repeats simulate-and-refit so a recovery claim has a bias and a spread
instead of one lucky seed; `suggest_model()` reads the shape of the data and names the
term each shape asks for; `benchmark_fit()` refuses to time anything fewer than three
times.

## 9. What the package does NOT do silently

These choices live in the errors, on purpose:

- a parent that is cited but has no line of its own is an ERROR, not a founder;
- a genotype outside 0, 1, 2 and NA is refused (NA is imputed to the mean; zero is a
  valid genotype and is never used as a missing-value code);
- a pen mate not in the pedigree is an ERROR, not a silent discard;
- an INADMISSIBLE theta halts the fit, it never turns into a result;
- a fit that did not converge SAYS so in the print, in the message and in the object.
