---
title: "Hands-on: every function, step by step"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hands-on: every function, step by step}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2026)
```

Every exported function of the package, exercised in order of a real workflow, on data
simulated in the document itself. Each chunk runs at build time, so this walkthrough
cannot drift away from the code. The companion vignette ("from data to breeding value")
tells the story; this one is the tool bench.

## Step 0 — simulate a population to work on

`simulate_breeding()` gene-drops haplotypes, so the genotypes it emits are consistent
with the pedigree it emits — one honest generator for every example.

```{r}
library(BreedingR)
s <- simulate_breeding(n_founders = 40, n_generations = 2,
                       offspring_per_generation = 80, h2 = 0.4,
                       n_markers = 200, seed = 7)
str(s, max.level = 1)
```

## Step 1 — look before estimating

```{r}
describe(s$data, "y", classes = "cg", pedigree = s$pedigree)
suggest_model(s$data, "y", "id")
```

## Step 2 — the pedigree toolbox

```{r}
p <- pedigree(s$pedigree)      # topological order + Meuwissen-Luo inbreeding
head(p, 3)
ai <- a_inverse(s$pedigree)    # Henderson's sparse A^-1, as triplets i/j/x/n
str(ai[c("i", "j", "x", "n")])
```

`a22_inverse()` takes 1-based positions in the topological order of `pedigree()`:

```{r}
gid <- s$genotypes$ids[141:200]
a22i <- a22_inverse(s$pedigree, geno = match(gid, p$id))
dim(a22i)
```

## Step 3 — fit the base animal model, read everything out of it

```{r}
fit <- model(y ~ cg + animal(id), s$data, s$pedigree)
fit
summary(fit)
coef(fit)
h2 <- fit$theta[["var(animal)"]] / sum(fit$theta)
round(h2, 3)                                # simulated truth: 0.4
```

```{r}
e <- ebv(fit)                               # named vector, all pedigree animals
acc <- accuracy(fit, s$pedigree)            # with the (1+F) in the denominator
round(cor(e[names(s$tbv)], s$tbv), 2)       # against the simulator's own truth
summary(acc)
```

## Step 4 — the formula markers, one by one

Repeated records ask for a permanent environment; `random()` is the iid effect for
litter or batch; a missing-value code leaves the likelihood via `missing_code=`:

```{r}
d3 <- s$data[rep(1:nrow(s$data), each = 2), ]
d3$y <- d3$y + rnorm(nrow(d3), 0, 0.5)
d3$y[1] <- -999
d3$lot <- sample(c("l1", "l2", "l3"), nrow(d3), TRUE)
f4 <- model(y ~ cg + random(lot) + animal(id) + pe(id), d3, s$pedigree,
            missing_code = -999)
round(f4$theta, 3)
```

Willham's full maternal model: direct and maternal genetic share ONE covariance matrix
via `group=`, and the two permanent environments disambiguate by column:

```{r}
d5 <- d3
d5$dam <- s$pedigree$dam[match(d5$id, s$pedigree$id)]
d5 <- d5[d5$dam != "0", ]
f5 <- model(y ~ cg + animal(id, group = "g") + maternal(dam, group = "g") +
              pe(id) + pe(dam), d5, s$pedigree, missing_code = -999)
names(f5$theta)
```

The reaction norm is `rn()` over a basis you build with `legendre()`; `h2_curve()` and
`plot()` then read h2 along the gradient:

```{r}
d6 <- s$data
d6$thi <- runif(nrow(d6), 55, 80)
d6 <- cbind(d6, legendre(d6$thi, order = 1, limits = c(55, 80)))
f6 <- model(y ~ cg + rn(id, base = c("phi0", "phi1")) + pe(id), d6, s$pedigree)
head(h2_curve(f6, limits = c(55, 80), points = 5))
plot(f6, limits = c(55, 80))
```

The indirect genetic effect (Muir & Schinckel's associative model) marks the PEN MATES
in the incidence; direct and indirect share a group so their correlation is estimated:

```{r}
d7 <- s$data
d7$penc <- rep(sprintf("p%03d", 1:50), length.out = nrow(d7))
f7 <- model(y ~ cg + animal(id, group = "g") + indirect(id, pen = "penc", group = "g"),
            d7, s$pedigree)
round(f7$theta, 3)
```

A sire model uses `sire()` when only males carry the genetic term. Unknown-parent
groups enter as metafounders (Legarra): a parent label with no line of its own,
declared with its gamma:

```{r}
ped8 <- s$pedigree
ped8$sire[ped8$sire == "0"] <- "L1"
ped8$dam[ped8$dam == "0"] <- "L2"
f8 <- model(y ~ cg + animal(id), s$data, ped8,
            metafounders = c("L1", "L2"), gamma = c(0.7, 0.7))
f8$converged
```

## Step 5 — genomics: files, QC, and the single step

`read_plink()` reads the binary trio (or a .raw); here we write a tiny .bed by hand so
the example carries its own truth:

```{r}
pref <- file.path(tempdir(), "toy")
codes <- c(`2` = 0L, `1` = 2L, `0` = 3L)                 # 2 bits per genotype
pack <- function(g) { v <- codes[as.character(g)]
  sum(v * 4L^(seq_along(v) - 1L)) }
con <- file(paste0(pref, ".bed"), "wb")
writeBin(as.raw(c(0x6c, 0x1b, 0x01, pack(c(2, 1, 0, 2)), pack(c(0, 0, 1, 1)))), con)
close(con)
writeLines(sprintf("F%d I%d 0 0 0 -9", 1:4, 1:4), paste0(pref, ".fam"))
writeLines(sprintf("1 SNP%d 0 %d A B", 1:2, 1:2), paste0(pref, ".bim"))
read_plink(pref)$m
```

`qc_genotypes()` filters call rate, MAF and (optionally) HWE, and reports what each
filter removed — silent filtering is how panels lose the markers that mattered:

```{r}
q <- qc_genotypes(s$genotypes$m, min_call_rate = 0.9, min_maf = 0.05)
c(kept = ncol(q$m), by_call = q$n_call, by_maf = q$n_maf)
```

The single step is an argument, not a function: `genotypes=` upgrades A to H. The
inverse of G* is exact by default, `apy_core=` uses a global core, `vecchia_k=` gives
each animal its own conditioning set:

```{r}
gen <- list(ids = gid, m = s$genotypes$m[141:200, ])
fss <- model(y ~ cg + animal(id), s$data, s$pedigree, genotypes = gen)
fap <- model(y ~ cg + animal(id), s$data, s$pedigree, genotypes = gen,
             apy_core = gid[1:30])
fve <- model(y ~ cg + animal(id), s$data, s$pedigree, genotypes = gen,
             vecchia_k = 25)
round(c(exact = cor(ebv(fss, "animal"), ebv(fap, "animal")),
        vecchia = cor(ebv(fss, "animal"), ebv(fve, "animal"))), 3)
```

The standalone inverses expose the same two approximations with their diagnostics
(the Mendelian/conditional residual of each animal):

```{r}
ap <- apy_inverse(gen$m, core = 1:30)
ve <- vecchia_inverse(gen$m, k = 25)
c(min_apy_res = round(min(ap$mendeliano), 3),
  min_vecchia_res = round(min(ve$mendeliano[-1]), 3))
```

Marker effects come out of a single-step fit by backsolving, or directly by fitting the
markers as equations — `snp_blup()`, the path that never builds G (theta is GIVEN, the
routine-evaluation flow):

```{r}
a_hat <- snp_effects(fss, s$pedigree, gen)
fsnp <- snp_blup(y ~ cg + animal(id), s$data, s$pedigree, gen,
                 theta = unname(fss$theta))
round(cor(unclass(a_hat), fsnp$g[!is.na(fsnp$g)]), 3)
```

## Step 6 — multi-trait

`cbind()` on the left, full R0, missingness handled record by record; `rg()` reads the
genetic correlation:

```{r}
dmt <- s$data
dmt$y2 <- 5 + 0.5 * s$tbv[dmt$id] + rnorm(nrow(dmt), 0, 0.8)
fmt <- model_mt(cbind(y, y2) ~ cg + animal(id), dmt, s$pedigree)
round(rg(fmt, "animal", "y", "y2"), 2)
head(ebv(fmt, "animal", trait = "y2"), 3)
```

## Step 7 — longitudinal data with a serially correlated residual

`model_ar1()` needs `subject=` and `time=`; with continuous time it is CAR(1). The
component `rho(residual)` is the reason this fitter exists:

```{r}
nlong <- 5
dar <- s$data[rep(1:nrow(s$data), each = nlong), ]
dar$day <- rep(1:nlong, nrow(s$data))
eps <- replicate(nrow(s$data), as.numeric(arima.sim(list(ar = 0.5), nlong, sd = 0.6)))
dar$y <- dar$y + c(eps)
far <- model_ar1(y ~ cg + animal(id) + pe(id), dar, s$pedigree,
                 subject = "id", time = "day")
round(far$theta[["rho(residual)"]], 2)
```

## Step 8 — the Bayesian half

`gibbs()` samples the SAME models; `ess()` and `geweke_z()` are the chain diagnostics
(also computed automatically on the result). With `theta_fixed=` it samples only the
locations — the mode whose posterior mean must be the BLUP:

```{r}
gb <- gibbs(y ~ cg + animal(id), s$data, s$pedigree,
            n_iter = 2000, burnin = 400, thin = 4)
print(gb)
round(c(ess = ess(gb$samples[, "var(animal)"]),
        geweke = geweke_z(gb$samples[, "var(animal)"])), 2)
```

## Step 9 — selection decisions

```{r}
idx <- selection_index(list(y = ebv(fit, "animal"),
                            y2 = ebv(fmt, "animal", trait = "y2")),
                       weights = c(y = 2, y2 = 1))
head(idx, 3)
drift <- rank_drift(ebv(fit, "animal"), ebv(fss, "animal"), top = 50)
c(spearman = round(drift$spearman, 3), retention = drift$retention)
```

## Step 10 — the environmental axis

```{r}
round(thi(c(22, 30), c(60, 85)), 1)          # NRC 1971
heat_load(c(60, 70, 76), threshold = 68)     # zero in comfort, linear above
```

## Step 11 — selection signatures

`fst()` needs two or more groups; `roh()` finds runs of homozygosity — here with a
planted 60-marker run, which comes back as an island:

```{r}
s2 <- simulate_breeding(n_founders = 40, n_generations = 1,
                        offspring_per_generation = 40, h2 = 0.4,
                        n_markers = 200, seed = 99)
sc <- fst(rbind(s$genotypes$m, s2$genotypes$m),
          groups = rep(c("A", "B"), c(nrow(s$genotypes$m), nrow(s2$genotypes$m))))
summary(sc$fst)

mroh <- s$genotypes$m[1:30, ]
mroh[1:10, 51:110] <- 2                       # the planted run
rr <- roh(mroh, min_snp = 30)
c(F_ROH_planted = round(mean(rr$animal$f_roh[1:10]), 2),
  island_peak = max(rr$marker$freq_roh))
```

## Step 12 — studies, timing, and the exposed internals

`mc_study()` repeats simulate-and-refit (cold starts, by the package's own rule);
`benchmark_fit()` refuses fewer than three replicates and checks they returned
identical numbers:

```{r}
m <- mc_study(n_rep = 3, h2 = 0.4, seed = 100, n_founders = 30,
              n_generations = 1, offspring_per_generation = 40)
round(c(mean_h2 = mean(m$h2_est), bias = attr(m, "bias")), 3)

b <- benchmark_fit(function() model(y ~ cg + animal(id), s$data, s$pedigree), reps = 3)
b
```

The internals are exported for the gates, and they double as sparse tools: the same
triplet list `i/j/x/n` that `a_inverse()` returns feeds all three:

```{r}
ch <- sparse_chol(ai)                        # minimum degree + symbolic + factor
x1 <- sparse_solve(ai, rep(1, ai$n))         # A^-1 %*% 1
si <- selected_inverse(ai)                   # A at the factor's pattern (PEV reads this)
round(c(logdet = ch$logdet, x1_first = x1[1]), 3)
inv_pd(matrix(c(4, 1, 1, 3), 2))             # dense positive-definite inverse
br_version()
```

And `eval_internal()` (with `_mt` and `_ar1` twins) evaluates -2logL by the TWO
independent routes at any theta — the identity every gate in the package leans on:

```{r}
ev <- eval_internal(y ~ cg + animal(id), s$data, s$pedigree, theta = c(0.4, 0.6))
round(abs(ev$neg2logl - ev$neg2logl_V), 10)
```

## The rule behind the tour

A new model is almost never a new function — it is a marker inside the formula of one
engine. A sibling fitter exists only when the mathematics changes: `model_mt()` (the
residual becomes R0), `model_ar1()` (the residual gains serial correlation), `gibbs()`
(the algorithm samples instead of maximizing), `snp_blup()` (a fixed-components solver
with markers as equations). Everything else orbits the fit.
