---
title: "Hands-on: every function, step by step"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hands-on: every function, step by step}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2026)
```

Forty-seven of the fifty-three exported functions, exercised in the order of a real
workflow, on data simulated in the document itself; the few left out — the inference
helpers `se_function()` and `profile_theta()`, `qc_phenotypes()`,
`competition_strength()`, and the `_mt`/`_ar1` twins of `eval_internal()` — are
documented on their own help pages and exercised by the test suite. Each chunk runs at build time, so this walkthrough
cannot drift away from the code. The companion vignette ("from data to breeding value")
tells the story; this one is the tool bench.

## Step 0 — simulate a population to work on

`simulate_breeding()` gene-drops haplotypes, so the genotypes it emits are consistent
with the pedigree it emits — one honest generator for every example.

```{r}
library(BreedingR)
s <- simulate_breeding(n_founders = 40, n_generations = 2,
                       offspring_per_generation = 80, h2 = 0.4,
                       n_markers = 200, seed = 7)
str(s, max.level = 1)
```

## Step 1 — look before estimating

```{r}
describe(s$data, "y", classes = "cg", pedigree = s$pedigree)
suggest_model(s$data, "y", "id")
```

## Step 2 — the pedigree toolbox

```{r}
p <- pedigree(s$pedigree)      # topological order + Meuwissen-Luo (1992) inbreeding
head(p, 3)
ai <- a_inverse(s$pedigree)    # Henderson's (1976) sparse A^-1, as triplets i/j/x/n
str(ai[c("i", "j", "x", "n")])
```

`a22_inverse()` takes 1-based positions in the topological order of `pedigree()`:

```{r}
gid <- s$genotypes$ids[141:200]
a22i <- a22_inverse(s$pedigree, geno = match(gid, p$id))
dim(a22i)
```

## Step 3 — fit the base animal model, read everything out of it

```{r}
fit <- model(y ~ cg + animal(id), s$data, s$pedigree)
fit
summary(fit)
coef(fit)
v <- fit$theta[grepl("^var[(]", names(fit$theta))]   # never sum a covariance or a rho
h2 <- v[["var(animal)"]] / sum(v)                    # both halves picked BY NAME
round(h2, 3)                                # simulated truth: 0.4
```

The fixed effects travel in the same solution vector, named `term=level`. The
parametrization is the implicit intercept with dropped columns at zero, so a table from
another convention (the book's, BLUPF90's) agrees only on CONTRASTS between levels:

```{r}
coef(fit, effects = "fixed")                # the same vector as fit$b
```

```{r}
e <- ebv(fit)                               # named vector, all pedigree animals
acc <- accuracy(fit, s$pedigree)            # with the (1+F) in the denominator
round(cor(e[names(s$tbv)], s$tbv), 2)       # against the simulator's own truth
summary(acc)
```

## Step 4 — the formula markers, one by one

Repeated records ask for a permanent environment; `random()` is the iid effect for
litter or batch; a missing-value code leaves the likelihood via `missing_code=`:

```{r}
d3 <- s$data[rep(1:nrow(s$data), each = 2), ]
d3$y <- d3$y + rnorm(nrow(d3), 0, 0.5)
d3$y[1] <- -999
d3$lot <- sample(c("l1", "l2", "l3"), nrow(d3), TRUE)
f4 <- model(y ~ cg + random(lot) + animal(id) + pe(id), d3, s$pedigree,
            missing_code = -999)
round(f4$theta, 3)
```

The full maternal model of Mrode and Pocrnic (2023, Eqn 8.1): direct and maternal genetic share
ONE covariance matrix via `group=`, and there is ONE permanent environment, the dam's,
which carries her non-additive maternal genetics as well. `d5` has two records per
animal, so a second `pe()` on the animal itself is identifiable here too; on one record
per animal it would BE the residual. Two `pe()` in one model must be NAMED:

```{r}
d5 <- d3
d5$dam <- s$pedigree$dam[match(d5$id, s$pedigree$id)]
d5 <- d5[d5$dam != "0", ]
f5 <- model(y ~ cg + animal(id, group = "g") + maternal(dam, group = "g") +
              pe(id, nome = "pe_animal") + pe(dam, nome = "pe_dam"),
            d5, s$pedigree, missing_code = -999)
names(f5$theta)
```

The reaction norm is `rn()` over a basis you build with `legendre()`; `h2_curve()` and
`plot()` then read h2 along the gradient:

```{r}
d6 <- s$data
d6$thi <- runif(nrow(d6), 55, 80)
d6 <- cbind(d6, legendre(d6$thi, order = 1, limits = c(55, 80)))
f6 <- model(y ~ cg + rn(id, base = c("phi0", "phi1")) + pe(id), d6, s$pedigree)
head(h2_curve(f6, limits = c(55, 80), points = 5))
plot(f6, limits = c(55, 80))
```

The indirect genetic effect (the associative model of Muir & Schinckel, 2002) marks the PEN MATES
in the incidence; direct and indirect share a group so their correlation is estimated:

```{r}
d7 <- s$data
d7$penc <- rep(sprintf("p%03d", 1:50), length.out = nrow(d7))
f7 <- model(y ~ cg + animal(id, group = "g") + indirect(id, pen = "penc", group = "g"),
            d7, s$pedigree)
round(f7$theta, 3)
```

A sire model uses `sire()` when only males carry the genetic term. Unknown-parent
groups enter as metafounders (Legarra et al., 2015): a parent label with no line of its own,
declared with its gamma:

```{r}
ped8 <- s$pedigree
ped8$sire[ped8$sire == "0"] <- "L1"
ped8$dam[ped8$dam == "0"] <- "L2"
f8 <- model(y ~ cg + animal(id), s$data, ped8,
            metafounders = c("L1", "L2"), gamma = c(0.7, 0.7))
f8$converged
```

## Step 5 — genomics: files, QC, and the single step

`read_plink()` reads the binary trio (or a .raw); here we write a tiny .bed by hand so
the example carries its own truth:

```{r}
pref <- file.path(tempdir(), "toy")
codes <- c(`2` = 0L, `1` = 2L, `0` = 3L)                 # 2 bits per genotype
pack <- function(g) { v <- codes[as.character(g)]
  sum(v * 4L^(seq_along(v) - 1L)) }
con <- file(paste0(pref, ".bed"), "wb")
writeBin(as.raw(c(0x6c, 0x1b, 0x01, pack(c(2, 1, 0, 2)), pack(c(0, 0, 1, 1)))), con)
close(con)
writeLines(sprintf("F%d I%d 0 0 0 -9", 1:4, 1:4), paste0(pref, ".fam"))
writeLines(sprintf("1 SNP%d 0 %d A B", 1:2, 1:2), paste0(pref, ".bim"))
read_plink(pref)$m
```

`qc_genotypes()` filters call rate, MAF and (optionally) HWE, and reports what each
filter removed — silent filtering is how panels lose the markers that mattered:

```{r}
q <- qc_genotypes(s$genotypes$m, min_call_rate = 0.9, min_maf = 0.05)
c(kept = ncol(q$m), by_call = q$n_call, by_maf = q$n_maf)
```

The single step is an argument, not a function: `genotypes=` upgrades A to H. The
inverse of G* is exact by default, `apy_core=` uses a global core, `vecchia_k=` gives
each animal its own conditioning set:

```{r}
gen <- list(ids = gid, m = s$genotypes$m[141:200, ])
fss <- model(y ~ cg + animal(id), s$data, s$pedigree, genotypes = gen)
fap <- model(y ~ cg + animal(id), s$data, s$pedigree, genotypes = gen,
             apy_core = gid[1:30])
fve <- model(y ~ cg + animal(id), s$data, s$pedigree, genotypes = gen,
             vecchia_k = 25)
round(c(exact = cor(ebv(fss, "animal"), ebv(fap, "animal")),
        vecchia = cor(ebv(fss, "animal"), ebv(fve, "animal"))), 3)
```

The standalone inverses expose the same two approximations with their diagnostics
(the Mendelian/conditional residual of each animal):

```{r}
ap <- apy_inverse(gen$m, core = 1:30)
ve <- vecchia_inverse(gen$m, k = 25)
c(min_apy_res = round(min(ap$mendeliano), 3),
  min_vecchia_res = round(min(ve$mendeliano[-1]), 3))
```

Marker effects come out of a single-step fit by backsolving, or directly by fitting the
markers as equations — `snp_blup()`, the path that never builds G (theta is GIVEN, the
routine-evaluation flow):

```{r}
a_hat <- snp_effects(fss, s$pedigree, gen)
fsnp <- snp_blup(y ~ cg + animal(id), s$data, s$pedigree, gen,
                 theta = unname(fss$theta))
round(cor(unclass(a_hat), fsnp$g[!is.na(fsnp$g)]), 3)
```

## Step 6 — multi-trait

`cbind()` on the left, full R0, missingness handled record by record; `rg()` reads the
genetic correlation:

```{r}
dmt <- s$data
dmt$y2 <- 5 + 0.5 * s$tbv[dmt$id] + rnorm(nrow(dmt), 0, 0.8)
fmt <- model_mt(cbind(y, y2) ~ cg + animal(id), dmt, s$pedigree)
round(rg(fmt, "animal", "y", "y2"), 2)
head(ebv(fmt, "animal", trait = "y2"), 3)
```

## Step 7 — longitudinal data with a serially correlated residual

`model_ar1()` needs `subject=` and `time=`; with continuous time it is CAR(1). The
component `rho(residual)` is the reason this fitter exists:

```{r}
nlong <- 5
dar <- s$data[rep(1:nrow(s$data), each = nlong), ]
dar$day <- rep(1:nlong, nrow(s$data))
eps <- replicate(nrow(s$data), as.numeric(arima.sim(list(ar = 0.5), nlong, sd = 0.6)))
dar$y <- dar$y + c(eps)
far <- model_ar1(y ~ cg + animal(id) + pe(id), dar, s$pedigree,
                 subject = "id", time = "day")
round(far$theta[["rho(residual)"]], 2)
```

## Step 8 — the Bayesian half

`gibbs()` samples the SAME models; `ess()` and `geweke_z()` are the chain diagnostics
(also computed automatically on the result). With `theta_fixed=` it samples only the
locations — the mode whose posterior mean must be the BLUP:

```{r}
gb <- gibbs(y ~ cg + animal(id), s$data, s$pedigree,
            n_iter = 2000, burnin = 400, thin = 4)
print(gb)
round(c(ess = ess(gb$samples[, "var(animal)"]),
        geweke = geweke_z(gb$samples[, "var(animal)"])), 2)
```

## Step 9 — selection decisions

```{r}
idx <- selection_index(list(y = ebv(fit, "animal"),
                            y2 = ebv(fmt, "animal", trait = "y2")),
                       weights = c(y = 2, y2 = 1))
head(idx, 3)
drift <- rank_drift(ebv(fit, "animal"), ebv(fss, "animal"), top = 50)
c(spearman = round(drift$spearman, 3), retention = drift$retention)
```

## Step 10 — the environmental axis

```{r}
round(thi(c(22, 30), c(60, 85)), 1)          # NRC 1971
heat_load(c(60, 70, 76), threshold = 68)     # zero in comfort, linear above
```

## Step 11 — selection signatures

`fst()` needs two or more groups; `roh()` finds runs of homozygosity — here with a
planted 60-marker run, which comes back as an island:

```{r}
s2 <- simulate_breeding(n_founders = 40, n_generations = 1,
                        offspring_per_generation = 40, h2 = 0.4,
                        n_markers = 200, seed = 99)
sc <- fst(rbind(s$genotypes$m, s2$genotypes$m),
          groups = rep(c("A", "B"), c(nrow(s$genotypes$m), nrow(s2$genotypes$m))))
summary(sc$fst)

mroh <- s$genotypes$m[1:30, ]
mroh[1:10, 51:110] <- 2                       # the planted run
rr <- roh(mroh, min_snp = 30)
c(F_ROH_planted = round(mean(rr$animal$f_roh[1:10]), 2),
  island_peak = max(rr$marker$freq_roh))
```

## Step 12 — studies, timing, and the exposed internals

`mc_study()` repeats simulate-and-refit (cold starts, by the package's own rule);
`benchmark_fit()` refuses fewer than three replicates and checks they returned
identical numbers:

```{r}
m <- mc_study(n_rep = 3, h2 = 0.4, seed = 100, n_founders = 30,
              n_generations = 1, offspring_per_generation = 40)
round(c(mean_h2 = mean(m$h2_est), bias = attr(m, "bias")), 3)

b <- benchmark_fit(function() model(y ~ cg + animal(id), s$data, s$pedigree), reps = 3)
b
```

The internals are exported for the gates, and they double as sparse tools: the same
triplet list `i/j/x/n` that `a_inverse()` returns feeds all three:

```{r}
ch <- sparse_chol(ai)                        # minimum degree (George & Liu 1989) + symbolic + factor
x1 <- sparse_solve(ai, rep(1, ai$n))         # SOLVES A^-1 x = 1, so x = A %*% 1: the
                                             # row sums of A, without ever forming A
si <- selected_inverse(ai)                   # A at the factor's pattern (PEV reads this)
round(c(logdet = ch$logdet, x1_first = x1[1]), 3)
inv_pd(matrix(c(4, 1, 1, 3), 2))             # dense positive-definite inverse
br_version()
```

And `eval_internal()` (with `_mt` and `_ar1` twins) evaluates -2logL by the TWO
independent routes at any theta — the identity every gate in the package leans on:

```{r}
ev <- eval_internal(y ~ cg + animal(id), s$data, s$pedigree, theta = c(0.4, 0.6))
round(abs(ev$neg2logl - ev$neg2logl_V), 10)
```

## Step 13 — declared covariances: dominance, epistasis, multibreed

`kernel(id, K=)` is the marker for a random term whose covariance matrix is DECLARED
rather than derived from the pedigree or the markers inside the fit. The constructors
build the K: `dominance_matrix()` is Cockerham's D from the pedigree, and here the data
gets a planted dominance deviation (variance 0.3) so the fit has something true to find.
Repeated records keep the dominance separable from the residual — with one record per
animal the two are confounded up to the sib covariances, and the fit runs to a boundary
and says so:

```{r}
D <- dominance_matrix(s$pedigree)
set.seed(11)
dv <- drop(chol(D + diag(1e-8, nrow(D))) %*% rnorm(nrow(D))) * sqrt(0.3)
names(dv) <- rownames(D)
dk <- s$data[rep(1:nrow(s$data), each = 2), ]
dk$y <- 10 + s$tbv[dk$id] + dv[dk$id] + rnorm(nrow(dk), 0, sqrt(0.6))
fd <- model(y ~ cg + animal(id) + kernel(id, K = D), dk, s$pedigree)
round(fd$theta, 3)                          # planted: 0.3 kernel, 0.6 residual
```

The genomic versions build K from markers: `g_matrix()` (raw VanRaden G, the book's —
no blend, no A22 adjustment), `g_dominance()` (Vitezica's genomic D), `g_epistasis()`
(the Hadamard square with its diagonal averaged to 1), and `genomic_inbreeding()`
(f = 1 - het/N, fitted as a `cov()` and read back from `fit$b` as the inbreeding
depression):

```{r}
G  <- g_matrix(gen)
GD <- g_dominance(gen)
GE <- g_epistasis(G)
fin <- genomic_inbreeding(gen)
round(c(diag_G = mean(diag(G)), diag_D = mean(diag(GD)),
        diag_AA = mean(diag(GE)), mean_f = mean(fin)), 3)
```

`partial_a()` splits A by breed origin (Garcia-Cortes and Toro's partial matrices): one
matrix per founder breed, one per segregating pair, plus the breed-fraction table `f`
and the segregation coefficients `h`. A row of K that is entirely zero declares a level
with no contribution — a purebred gets no equation in the other breed's term — and each
matrix goes into its own `kernel()`:

```{r}
pedm <- data.frame(animal = c("1", "2", "3", "4", "5"),
                   sire   = c("0", "0", "1", "1", "3"),
                   dam    = c("0", "0", "2", "2", "4"))
pa <- partial_a(pedm, breed = c("1" = "A", "2" = "B"))
pa$f                                        # 3 and 4 are F1; 5 is F2
round(pa$K[["A"]], 2)                       # the A-partial: animal 2 contributes nothing
pa$h                                        # only the F2 segregates
```

## Step 14 — a categorical trait on the liability scale

`model_threshold()` fits an ordered categorical trait by probit on an underlying
liability (Gianola and Foulley), residual FIXED at 1 for identifiability, thresholds in
place of the intercept, components GIVEN via `start=`. `predict()` returns the
per-category probabilities:

```{r}
dt <- s$data
dt$cat <- as.integer(cut(dt$y, quantile(dt$y, c(0, 0.6, 0.85, 1)),
                         include.lowest = TRUE))
ft <- model_threshold(cat ~ cg + animal(id), dt, s$pedigree, start = 0.35,
                      verbose = FALSE)
round(ft$thresholds, 3)
head(ebv(ft), 3)                            # on the liability scale
round(predict(ft, dt[1:2, ]), 3)            # P(category) per record
```

`h2_observed()` and `h2_liability()` are the Dempster and Lerner conversion between the
0/1 observed scale and the liability scale — at incidence 0.234 half the heritability
is scale, not biology:

```{r}
round(c(observed = h2_observed(0.1781, 0.234),
        liability = h2_liability(0.0934, 0.234)), 4)
```

## Step 15 — time until failure, with censoring

`model_survival()` fits the Weibull frailty model (Kachman); `censor=` is mandatory
because a censored record is a LOWER BOUND on the failure time, not a missing value.
This chunk is Example 16.1 of Mrode and Pocrnic — length of productive life of 12 cows,
4 censored — and the numbers it prints are the book's p.291-292:

```{r}
cows <- data.frame(
  cow  = as.character(8:19),
  herd = as.character(c(1,1,1,1,1,1,2,2,2,2,2,2)),
  ysp  = as.character(c(3,4,1,2,3,1,4,1,2,3,4,2)),
  code = c(0,1,0,1,1,1,1,1,0,1,0,1),        # 0 = censored
  lpl  = c(40,47,22,28,50,33,49,29,23,37,35,30))
peds <- data.frame(
  animal = as.character(1:19),
  sire = c(rep("0", 7), "1","1","4","4","5","5","1","1","5","5","4","4"),
  dam  = c(rep("0", 7), "2","3","2","9","3","8","6","7","14","6","7","3"))
fsu <- model_survival(lpl ~ herd + ysp + animal(cow), cows, peds, censor = "code",
                      rho = 1, lambda = 1, sigma2 = 0.4, verbose = FALSE)
round(coef(fsu, "fixed"), 3)                # herd=2 -1.631, ysp -2.346/-3.149/-2.982
round(exp(ebv(fsu)[["1"]]), 3)              # relative risk of animal 1: 0.459
round(predict(fsu, data.frame(herd = "1", ysp = "4", cow = "1"),
              time = 40, type = "survival"), 3)   # S(40) = 0.394
```

## The rule behind the tour

A new model is almost never a new function — it is a marker inside the formula of one
engine, `kernel()` being the marker that takes ANY declared covariance. A sibling
fitter exists only when the mathematics changes: `model_mt()` (the residual becomes
R0), `model_ar1()` (the residual gains serial correlation), `gibbs()` (the algorithm
samples instead of maximizing), `snp_blup()` (a fixed-components solver with markers as
equations), `model_threshold()` (the residual stops being Gaussian with a free
variance), `model_survival()` (there is no residual at all — the randomness is the
Weibull the linear predictor scales). Everything else orbits the fit.
