---
title: "Theory and practice: a genetic evaluation from end to end"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theory and practice: a genetic evaluation from end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2026)
library(BreedingR)
```

This document follows a genetic evaluation in the order it actually happens, and at each
step it says three things: what the matrices are and what they mean, what the algorithm
does with them, and which function runs it. The formulas are the ones the package
computes, not the textbook version they descend from — where the implementation makes a
choice, the choice is stated.

The data for every example is simulated here, so each result has its truth beside it.

```{r}
s <- simulate_breeding(n_founders = 60, n_generations = 3,
                       offspring_per_generation = 150, h2 = 0.4,
                       n_markers = 600, seed = 1)
str(s, max.level = 1)
```

## 1. The model

Every fit in this package is a linear mixed model,

```
y = X b + Z u + e,     u ~ N(0, D),     e ~ N(0, R)
```

where `y` holds the observations, `b` the fixed effects, `u` all the random effects
stacked, and `X` and `Z` their incidence matrices. What separates one model from another
is not the algebra but the shape of `D` and `R`: an animal model puts the pedigree
relationship matrix inside `D`, a repeatability model adds a second block, a multi-trait
model gives `R` a full covariance between traits, a reaction norm gives each animal a
vector of coefficients instead of a scalar. Most of the literature calls that covariance
`G`; this document reserves `G` for the genomic relationship matrix of section 6 and
writes `D` for the covariance of the random effects throughout.

The package writes all of that in one formula. Fixed terms are bare, `cov()` marks a
fixed covariate, and the marked terms are random: `animal()`, `maternal()`, `sire()`
carry a relationship matrix, `pe()` and `random()` are independent, `rn()` is a random
regression, `indirect()` is the associative effect of pen mates.

`pe()` is worth its full definition, because it is the term the repeatability model is
named after. It is the permanent environment of a subject with repeated records: what all
the records of that subject share and is NOT additive genetic, so the non-additive
genetic effects land there too (Mrode and Pocrnic, 2023, Eqn 5.1). The repeatability
itself is `(var(animal) + var(pe)) / total`, the two `share` cells of the printed table
added together.

**The unit of layout is the covariance group, not the term.** Two terms that share
`group =` go into ONE covariance matrix with the covariance between them estimated:

```
Var(u_g) = C_g (x) K_g
```

where `C_g` is the small covariance among the terms of the group and `K_g` is the
relationship matrix among their levels; `D` is the block diagonal of these, one block
per group. This is why direct-maternal, the reaction norm and the associative model have
no dedicated fitter — they are the same engine with a different incidence and the same
penalty.

## 2. The pedigree: A, F, and A inverse

The additive relationship matrix `A` holds twice the kinship between every pair of
animals. It is never built. What the equations need is its inverse, and Henderson (1976)
showed that the inverse has a closed form: for each animal `i` with Mendelian sampling
variance `d_i`,

```
A^-1 = sum_i (1/d_i) v_i v_i',    v_i = +1 at i, -1/2 at sire, -1/2 at dam
```

so `A^-1` is built by adding one small outer product per animal — sparse from the start,
never through a dense `A`. The Mendelian variance depends on how much of the parents'
ancestry is already shared:

```
d_i = 1/2  - 1/4 (F_sire + F_dam)   both parents known
d_i = 3/4  - 1/4 F_known            one parent known
d_i = 1                             neither
```

`F` is Wright's (1922) inbreeding coefficient, computed by the Meuwissen and Luo (1992) trace: for
animal `i`, walk the ancestors in decreasing index order accumulating `l_j` contributions,
and `F_i = sum_j l_j^2 d_j - 1`. A max-heap over the ancestors is what makes a single
pass correct — every descendant of an ancestor has a larger index and has therefore
already deposited its share. This trace and the earlier strategy of Quaas (1976) share
the point: the diagonal of `A` comes out of the decomposition while `A^-1` is being
built, instead of storing `A` to read it back. Mrode and Pocrnic (2023, ch. 3) carry both
rules through a small pedigree by hand, which is where an implementation gets checked
against printed numbers.

```{r}
p <- pedigree(s$pedigree)
head(p, 3)
ai <- a_inverse(s$pedigree)      # i, j, x, n: the lower triangle in triplets
c(animals = ai$n, nonzeros = length(ai$x), mean_F = round(mean(p$F), 4))
```

Two details in the implementation are worth knowing because they are silent when wrong.
A parent that is cited but has no line of its own is an error, not a new founder. And
when sire and dam are the same animal, the coefficients are merged BEFORE the outer
product, which puts `1/d` on that parent's diagonal; emitting the terms one by one into a
half-stored triangle would give `3/4 d^-1` and no error message.

**Metafounders.** When the base population is not one homogeneous unrelated pool —
different lines, different years, animals whose parents are simply unknown — Legarra et
al. (2015) replace the unknown parents with *metafounders*: virtual base animals with a
self-relationship `gamma` that measures how related that base actually was. The
implementation reduces to one line: a metafounder row is preloaded with
`F = gamma - 1` and `d = gamma`, and because every formula above reads only the parents'
`F`, the whole generalization propagates through the Mendelian variances, the trace and
`A^-1` with no special case. `gamma` must be strictly inside `(0, 2)`, and the two ends
are refused for different reasons. The upper one is Mendelian: the offspring of two copies
of the same metafounder has `d = 1/2 - 1/4 (2 (gamma - 1)) = 1 - gamma/2`, which vanishes
at `gamma = 2` — a base pool completely homozygous by descent, where every base animal is
the same animal. The lower one is that a base pool needs a positive self-relationship to
be a pool at all, and the package's error message names both. Zero is a limit with a
meaning rather than a degeneracy: `gamma -> 0` is complete heterozygosity by descent, an
unrelated base, and the recursion collapses onto the classical unknown parent of
Meuwissen and Luo (1992) (Legarra et al., 2015; Mrode and Pocrnic, 2023, sec. 3.8).

```{r}
ped_mf <- s$pedigree
ped_mf$sire[ped_mf$sire == "0"] <- "L1"
ped_mf$dam[ped_mf$dam == "0"] <- "L1"
p_mf <- pedigree(ped_mf, metafounders = "L1", gamma = 0.7)
round(mean(p_mf$F[p_mf$id %in% s$pedigree$id]), 4)   # base relatedness now shows up as F
```

## 3. The mixed model equations

Henderson's equations (Henderson, 1950) solve for `b` and `u` at once. Writing `W = [X Z]`, the package
assembles

```
C_s = W'W + blockdiag_g [ kron(s2e * C_g^-1, K_g^-1) ],     rhs = W'y
```

and solves `C_s sol = rhs`. Two conventions matter. The system is scaled by the residual
variance — the penalty carries `s2e * C_g^-1`, not `C_g^-1` — so the solution is the same
BLUP but the matrix is in units of `s2e`. And inside a term, the column of `Z` for
coefficient `a` and level `l` is `a * n_levels + l`: the **level varies fastest**.
Reversing that convention does not break anything visibly — the model still assembles and
converges, and estimates something else.

`W'W` is accumulated one record at a time. Each record touches a handful of columns and
contributes a small clique among them, so the cost is the sum of squared row-lengths, not
anything quadratic in the number of columns. The penalty is added afterwards into the
same triplet list, and the triplet-to-sparse conversion **sums duplicates** — which is
not a convenience but a requirement: in `A^-1` the mate-mate term arrives once per
offspring, and a build where the last write wins would silently discard part of the
matrix.

The restricted likelihood, in the form the MME give:

```
-2logL = (n - p) log s2e + log|D| + log|C_s| + (y'y - sol'rhs) / s2e
```

`D` is the covariance of the random effects of section 1, and it carries the same `s2e`
scaling as `C_s`: per group the code accumulates `nl log|C_g / s2e| + dim log|K_g|`. With
`weights =`, a `- sum_i log w_i` term joins the expression — the Jacobian of the weighted
rescaling, constant in theta, so it moves the value and not the optimum.

The package computes the same quantity a second way, through the dense
`V = Z D Z' + R`, and the tests demand the two agree. They share nothing but the model,
so agreement is evidence; a single path checked against itself would be none.

```{r}
d <- s$data
ev <- eval_internal(y ~ cg + animal(id), d, s$pedigree, theta = c(0.4, 0.6))
c(mme = ev$neg2logl, dense_V = ev$neg2logl_V, difference = abs(ev$neg2logl - ev$neg2logl_V))
```

## 4. Solving: sparse Cholesky and the selected inverse

`C_s` is factored as `L L'` by an up-looking sparse Cholesky, preceded by a minimum
degree ordering (George and Liu, 1989) that reduces fill-in and by a symbolic analysis that computes the pattern
of `L` once. The symbolic factorization is cached across REML iterations, which is sound
only because the pattern of `C_s` does not depend on the variance components: the
assembly pushes the same slots every time, explicit zeros included.

Prediction error variances need the diagonal of `C_s^-1`, and the full inverse of a
sparse matrix is dense — for a pedigree of any size, unaffordable. The identity of
Takahashi, Fagan and Chin (1973) gives exactly the elements of the inverse that sit on
the pattern of `L`, which is all the diagonal needs:

```{r}
ch <- sparse_chol(ai)
si <- selected_inverse(ai)
c(logdet = round(ch$logdet, 4), dense_block = ch$dense_block)
```

A position outside the pattern is not zero — it is unknown. The function returns only
what it computed.

## 5. Estimating the variance components

REML maximizes the likelihood of error contrasts instead of the likelihood of `y`. What
that removes is not a bias in `b` — the fixed effects come out as the same GLS solution
under either criterion — but the downward bias maximum likelihood carries IN THE VARIANCE
COMPONENTS, whose cause is that those fixed effects were estimated from the same records.
Mrode and Pocrnic (2023, ch. 17) write the restricted likelihood as three terms: a
weighted sum of squares of the residuals, `log|V|`, and `log|X'V^-1X|`, the last of them a
penalty for having estimated `b` and the one maximum likelihood leaves out (Patterson and
Thompson, 1971). The `n - p` in the exponent of section 3 is that same correction seen
from the side of the mixed model equations.

The package maximizes that restricted likelihood by average information REML (Gilmour,
Thompson and Cullis, 1995; the animal-breeding derivation is in Mrode and Pocrnic, 2023,
ch. 17), which needs three ingredients per iteration.

**The score.** For a covariance group with matrix `C_g`, define the quadratic of the
solution and the trace from the selected inverse,

```
Q[a,b] = u_a' K^-1 u_b
T[a,b] = s2e * tr( K^-1 [C_s^-1]_{a,b} )
M_g    = nl * C_g^-1 - C_g^-1 (Q + T) C_g^-1
```

and the derivative of `-2logL` with respect to the element `(i,j)` of `C_g` is `M_g[i,i]`
on the diagonal and `M_g[i,j] + M_g[j,i]` off it, because a symmetric element appears
twice. For the residual,

```
d(-2logL)/ds2e = (n - p - q + trpen)/s2e - e'e/s2e^2
```

where `q` is the number of random columns and `trpen` the trace of the penalty against
`C_s^-1`. The `- q + trpen` is not decoration: the random effects absorb degrees of
freedom too, and a version without it still converges — to the wrong place, reporting
convergence. That is the worst failure mode available, which is why the score is checked
against central finite differences on every parameter.

**The step.** The average information matrix is the average of the observed and expected
information, cheap because it needs no second derivatives, and the step is
`theta <- theta - (AI + lambda diag(AI))^-1 score`. This `lambda` is the
Levenberg-Marquardt damping (Levenberg, 1944; Marquardt, 1963), and it is a third
meaning of the letter: not the `s2e/s2a`
ratio that scales `K^-1` inside the mixed model equations of section 3, and not the ridge
that `apy_inverse()` calls `lambda`. The damping starts at 1e-3,
multiplies by ten whenever a candidate is rejected — inadmissible covariance, singular
damped matrix, or a likelihood that got worse — and divides by ten when one is accepted.
Before the AI loop, a few EM rounds run: EM (Dempster, Laird and Rubin, 1977) is
monotone and stays inside the cone, so it costs little and lands the search in the
right basin.

**Stopping.** Convergence is judged on the relative change of the components,
`sqrt(sum(dtheta^2)/sum(theta^2)) < tol`, never on an absolute threshold on the score,
whose norm grows with the number of records. Standard errors come from the inverse of the
AI matrix, `SE_k = sqrt(2 [AI^-1]_kk)`.

```{r}
fit <- model(y ~ cg + animal(id), d, s$pedigree)
fit
```

Two rules protect the result. A covariance that stops being positive-definite ends the
fit rather than being nudged back, and the starting point always comes from `var(y)`, so
no run inherits a neighbour's answer.

## 6. Genomics: from markers to H

**G.** VanRaden (2008) builds the genomic relationship matrix from centred marker
dosages. With `p_j` the allele frequency at marker `j` and `Z = M - 2p`,

```
G = Z Z' / (2 sum_j p_j (1 - p_j))
```

Missing genotypes are imputed with the marker mean and never with zero — zero is a valid
genotype, and confusing the two changes both the frequencies and every element of `G`
with no error to notice. Markers that are fixed or entirely missing are dropped and
counted.

**A22, and the trap.** The single step needs the pedigree relationships *among the
genotyped animals*, `A22`. That is the inverse of the genotyped block of the inverse, not
the genotyped block of `A^-1` — two matrices of the same shape, both symmetric, only one
correct. The package computes it by a sparse Schur complement,

```
A22^-1 = B22 - B21 B11^-1 B12
```

where `B11` is the non-genotyped block. `B11^-1` is never formed: one sparse
factorization and one triangular solve per genotyped animal do the same work.

**H.** Genomic and pedigree information describe the same animals on two different
scales, so `G` is first brought to the scale of `A22` by an affine adjustment that
matches both the mean diagonal and the mean off-diagonal, then blended:

```
b = (mean_diag(A22) - mean_off(A22)) / (mean_diag(G) - mean_off(G))
a = mean_off(A22) - b * mean_off(G)
G* = (1 - w) (a + b G) + w A22
```

The blend `w` (5 percent by default) keeps `G*` invertible and admits that markers are
not the whole truth. Then the single-step inverse of Aguilar et al. (2010) and
Christensen and Lund (2010) says the whole relationship matrix `H` has an inverse that
differs from `A^-1` only in the genotyped block:

```
H^-1 = A^-1 + [ 0        0            ]
              [ 0   G*^-1 - A22^-1    ]
```

This is the identity that makes the single step practical: nothing dense is needed for
the non-genotyped animals, and the correction is a small block added to a sparse matrix.
It is also self-checking — with `blend = 1`, `G*` becomes `A22` exactly, the correction
block vanishes, and `H^-1` must equal `A^-1` through the whole fit. That collapse is one
of the package's tests. Mrode and Pocrnic (2023, ch. 12) work the same identity through a
small herd, number by number.

```{r}
gen <- list(ids = s$genotypes$ids, m = s$genotypes$m)
f_ss <- model(y ~ cg + animal(id), d, s$pedigree, genotypes = gen)
f_id <- model(y ~ cg + animal(id), d, s$pedigree, genotypes = gen, blend = 1)
c(single_step = f_ss$neg2logl, blend_one = f_id$neg2logl, pedigree = fit$neg2logl)
```

**Inverting G\* without paying the cube.** Two approximations are available, and they are
the same idea at different granularity. APY (Misztal, Legarra and Aguilar, 2014; Mrode
and Pocrnic, 2023, sec. 12.4) splits the genotyped into a core and the rest; every
non-core animal is regressed on the core, and what is left is a scalar Mendelian residual:

```
P   = Gcc^-1 Gcn
m_j = g_jj - g_jc P_j
G_APY^-1 = [ Gcc^-1 + P Mnn^-1 P'   -P Mnn^-1 ]
           [ -Mnn^-1 P'              Mnn^-1   ]
```

The Vecchia recursion (Vecchia, 1988) generalizes it: each animal conditions on its own `k` nearest
relatives among the previous ones rather than on one shared core. Henderson's `A^-1` is
this same recursion with the parents as the conditioning set — exact there, because a
pedigree is Markovian. Schafer, Katzfuss and Owhadi (2021) prove the factor is optimal in
Kullback-Leibler divergence given the pattern, and that enlarging the sets never hurts.

An honest limit: both approximations cut arithmetic, not memory. `G*^-1` is a dense
matrix of the genotyped animals on all three paths.

```{r}
gid <- gen$ids
f_apy <- model(y ~ cg + animal(id), d, s$pedigree, genotypes = gen,
               apy_core = gid[seq(1, length(gid), by = 3)])
f_vec <- model(y ~ cg + animal(id), d, s$pedigree, genotypes = gen, vecchia_k = 60)
round(c(apy = cor(ebv(f_ss, "animal"), ebv(f_apy, "animal")),
        vecchia = cor(ebv(f_ss, "animal"), ebv(f_vec, "animal"))), 4)
```

**Markers as equations.** The other route never forms `G` at all. Liu et al. (2014) write
the equivalent model with the marker effects as unknowns, `u_g = Z g + p_g`, and the
joint precision turns the Schur term into blocks of `A^-1`, so most of the system is the
same sparse matrix the pedigree already gives. `snp_blup()` solves it by preconditioned
conjugate gradients, which need only matrix-vector products, and every application of
`A22^-1` uses the matrix-free identity of Masuda et al. (2017):

```
A22^-1 v = A^22 v - A^21 (A^11)^-1 A^12 v
```

one sparse factorization of the non-genotyped block, one triangular solve per iteration.
Nothing dense is ever built. The variance components are given rather than estimated:
this is the solver of routine evaluation, where components come from a periodic REML run.

## 7. What the formula can say

With the covariance group as the unit, several classical models are incidences rather
than fitters.

*Direct-maternal* (Willham, 1972): the animal's own genes and its dam's maternal ability
in one 2x2 matrix, with the direct-maternal covariance — usually negative, and the number
the model exists for.

*Reaction norm* (Kirkpatrick, Lofsvold and Bulmer, 1990; in animal models, Schaeffer
and Dekkers, 1994): each animal gets a vector of
coefficients on an environmental gradient through `rn(id, base = ...)`, with
`legendre()` building the orthogonal basis. The genetic variance becomes a function of
the environment, which `h2_curve()` reads.

*Indirect genetic effects* (Griffing, 1967; Muir and Schinckel, 2002; Bijma, Muir and
Van Arendonk, 2007): the incidence of a record marks the animal's PEN MATES, so a phenotype carries its
own genes plus the associative effects of those it lives with. Direct and indirect share
a group, so the fit returns `var(animal)`, `var(indirect)` and the covariance between
them, and the SIGN of that covariance is what separates heritable competition from
heritable co-operation.

What decides the response is not that covariance on its own but the total breeding value,
into which the group size enters twice:

```
TBV    = A_D + (n - 1) A_S
s2_TBV = s2_AD + 2 (n - 1) s_ADS + (n - 1)^2 s2_AS
t2     = s2_TBV / s2_p
```

Because `(n - 1)` multiplies the covariance and `(n - 1)^2` the associative variance, the
group size sets both the magnitude of the response and, under strong heritable
competition, its direction: the response need not follow the selection pressure (Bijma,
Muir and Van Arendonk, 2007; Mrode and Pocrnic, 2023, ch. 9). The package returns the
three components; combining them into `TBV` and `t2` is the caller's step, and it takes
the pen size.

The book fixes `n = 3`; real pens are unequal, and with a coefficient of 1 per mate the
social sum grows with `n_i - 1`. `indirect(id, pen = "pen", dilution = d)` scales every
mate's entry to `(n_i - 1)^(-d)`: `d = 0` (the default) is the book's sum, `d = 1` the
mate mean, and the regime in between is the dilution of Bijma (2010) — chosen
empirically by fitting a small grid of `d` and comparing `-2logL`, which stays
comparable because only the incidence changes. The residual inherits the same
heterogeneity, `var(e_i) = s2_ED + (n_i - 1) s2_ES`; `indirect_residual()` estimates
the ratio `s2_ES / s2_ED` by profile REML over weighted fits, and its documentation
says which of the two the data actually pins.

*Multi-trait*: `cbind()` on the left, a full residual covariance between traits, and
missingness handled by the pattern of each record rather than by dropping it.

*AR(1)/CAR(1) residual* (Wade and Quaas, 1993): for repeated measures along time, the residual correlation
decays as `rho^|dt|`. The inverse of that correlation matrix is tridiagonal — the
classical Markov result — so the model costs no more than the iid one, and `rho` comes
out with its standard error.

The textbook treatment of these models is Mrode and Pocrnic (2023): the repeatability
model in ch. 5, maternal traits in ch. 8, social interaction in ch. 9, random regression
in ch. 10.

```{r, eval = FALSE}
model(weight ~ cg + animal(id, group = "g") + maternal(dam, group = "g"), d, ped)
model(y ~ cg + rn(id, base = c("phi0", "phi1")) + pe(id), d2, ped)
model(y ~ cg + animal(id, group = "g") + indirect(id, pen = "pen", group = "g"), d, ped)
model_mt(cbind(t1, t2) ~ cg + animal(id), d, ped)
model_ar1(y ~ cg + animal(id) + pe(id), d, ped, subject = "id", time = "day")
```

## 8. Reading the evaluation

The solution of the MME gives the breeding values directly. Their reliability comes from
the prediction error variance, which is where the selected inverse pays off, and the
accuracy of an animal's EBV is

```
acc = sqrt( 1 - PEV / ((1 + F) sigma2_a) )
```

with the inbreeding coefficient in the denominator because an inbred animal has more
additive variance to predict: `var(u_i) = a_ii s2a = (1 + F_i) s2a` (Mrode and Pocrnic,
2023, ch. 3). The identity underneath, `PEV = var(a - a^) = C22 s2e`, is Henderson (1975).

One declared limit comes with it. `accuracy()` reads `F` from the PEDIGREE even when the
fit was single-step. For a genotyped animal the prior variance is the diagonal of `H`,
which in that block is the diagonal of `G*`, and that is not `1 + F_ped`: on the
population simulated at the top of this vignette, every animal genotyped, the two
diagonals differ by up to 0.18 and an individual accuracy by up to 0.067. The herd mean
does not move; the individual does. Read genomic accuracies with that in mind until the
function reads the `H` diagonal itself.

Heritability is read off `theta`, and both halves of the ratio have to be picked rather
than assumed. `theta` is a NAMED vector, and its first entry is the animal variance only
when the formula happens to put that term first — `random(lot) + animal(id)` puts
`var(random)` there. It also carries the covariances of a `group =` and the
`rho(residual)` of `model_ar1()`, which belong in no variance denominator. So: numerator
by name, denominator over the `var()` entries alone. That is what the `share` column of
the printed fit already computes.

The fixed effects come with the same solution vector: `fit$b`, named `term=level`, also
reachable as `coef(fit, effects = "fixed")`. The parametrization is the implicit
intercept, with linearly dependent columns dropped (`dropped_x`) and a dropped level
holding the zero solution — while the book's examples zero one level per factor and fit
no intercept. The two conventions agree only on CONTRASTS, differences between levels
of the same factor, never on the raw values; contrasts are invariant to the
parametrization, individual entries are not.

```{r}
e <- ebv(fit)
a <- accuracy(fit, s$pedigree)
v <- fit$theta[grepl("^var[(]", names(fit$theta))]     # the variances, and only them
round(c(cor_with_truth = cor(e[names(s$tbv)], s$tbv),
        mean_accuracy = mean(a), h2 = v[["var(animal)"]] / sum(v)), 3)
```

From there the decisions: `selection_index()` combines traits with economic weights, and
`rank_drift()` answers what an updated evaluation actually changed.

## 9. The Bayesian alternative

The same models can be sampled instead of maximized. `gibbs()` draws all locations in one
block through the same sparse Cholesky, then draws each covariance from its conjugate
inverse-Wishart and the residual from a scaled inverse chi-square, under reference priors.
Where REML gives a point and a standard error from the information matrix, the sampler
gives the whole posterior — worth the cost when a component sits near a boundary, where
the normal approximation behind the standard error is least trustworthy. The sampler
goes back to Geman and Geman (1984); Wang, Rutledge and Gianola (1993, 1994) brought it
to variance components in animal models, and Mrode and Pocrnic (2023, ch. 18) set out
the same sampler and the conjugate draws it uses.

```{r}
g <- gibbs(y ~ cg + animal(id), d, s$pedigree, n_iter = 2000, burnin = 400, thin = 4)
data.frame(component = colnames(g$samples),
           posterior_mean = round(unname(g$mean), 3),
           reml = round(unname(fit$theta), 3))
```

## 10. A declared covariance: dominance, epistasis, breed origin

Sections 2 and 6 derived the covariance of a random term from the pedigree or from the
markers. `kernel(id, K = )` removes the derivation: the term's covariance matrix is
DECLARED, symmetric positive-definite, its rownames naming the levels, and every animal
in `K` gets an equation whether or not it has a record. Nothing else changes — the same
`kron(C^-1, K^-1)` penalty, the same score, the same selected inverse — so any model
whose only novelty is the covariance structure is a constructor plus a marker, not a new
fitter (Mrode and Pocrnic, 2023, chs 13-14).

The constructors cover the classical cases. `dominance_matrix()` builds Cockerham's
(1954) D from the pedigree, `d_xy = 0.25 (a_sf a_dm + a_sd a_fm)` with diagonal 1 — the
covariance of the dominance deviations, fitted BESIDE the additive term so the two
partition the genetic variance. The genomic versions are `g_dominance()` (Vitezica,
Varona and Legarra, 2013) and `g_epistasis()`, the Hadamard square of `G` with its
diagonal averaged to one; `genomic_inbreeding()` reads `f = 1 - het/N` per animal, and
fitted as a `cov()` its coefficient in `fit$b` is the inbreeding depression.

For a multibreed pedigree, `partial_a()` splits `A` by breed origin (Garcia-Cortes and
Toro, 2006): one partial matrix per founder breed and one per segregating pair, built by
the tabular recursion with the breed contribution on the diagonal. A purebred
contributes nothing to the other breed's matrix, and its row there is entirely zero —
which `kernel()` reads as "no equation for this level, keep its records" — so each
matrix enters as its own term and the breed-specific solutions sum to the combined
breeding value.

```{r}
D <- dominance_matrix(s$pedigree)
f10 <- model(y ~ cg + animal(id) + kernel(id, K = D), d, s$pedigree)
round(f10$theta, 3)
```

Read that split with care. This data has ONE record per animal and was simulated with
no dominance, and `D` has a unit diagonal — so `var(kernel)` and `var(residual)` are
separated only by the sib covariances, a weakly identified partition, and the
likelihood is nearly flat between them. The estimate above is one point on that
plateau, not a discovery of dominance. The hands-on vignette plants a real dominance
signal on repeated records, where the partition is estimable and the fit recovers it.

## 11. The threshold model: a categorical trait on a liability

A calving score has ordered categories, not a variance. The threshold model puts a
Gaussian LIABILITY `l = Xb + Zu + e` underneath and maps it to the observed category
through thresholds: `y = j` when `t_{j-1} < l <= t_j`. Only signs and orderings of `l`
are observable, so the residual is FIXED at 1 (probit identifiability) and the
thresholds replace the intercept. `model_threshold()` fits it by Fisher scoring on the
Gianola and Foulley (1983) system, with the variance components on the liability scale
GIVEN via `start=` — the Gaussian REML of section 5 has no likelihood for a categorical
trait, and the package says so rather than pretending. `cbind(quant, binary)` fits the
joint quantitative-binary analysis of Foulley, Gianola and Thompson (1983); Mrode and
Pocrnic (2023, ch. 15) carry both through worked examples, which are the package's
gates.

The scale matters when comparing heritabilities: a 0/1 trait analysed linearly reports
an OBSERVED-scale h2, related to the liability one by the Dempster and Lerner (1950)
factor `z^2 / (p(1-p))` at incidence `p`.

```{r}
d11 <- d
d11$cat <- as.integer(cut(d11$y, quantile(d11$y, c(0, 0.6, 0.85, 1)),
                          include.lowest = TRUE))
f11 <- model_threshold(cat ~ cg + animal(id), d11, s$pedigree, start = 0.35,
                       verbose = FALSE)
round(f11$thresholds, 3)
round(h2_observed(0.35 / 1.35, 0.4), 3)    # what that liability h2 shows on a 0/1 scale
```

## 12. Time until failure

Longevity is a time, and part of it is unobserved: a cow still alive at culling day has
a LOWER BOUND on her length of productive life, not a missing value. The Weibull frailty
model (Kachman, 1999; Mrode and Pocrnic, 2023, ch. 16) writes the hazard as
`h(t) = h0(t) exp(x'b + z'u)` with a Weibull baseline; a complete record contributes its
density, a censored one its survival `S(t)`, and that difference is where the censoring
information lives — `model_survival()` therefore refuses to run without `censor=`.
Solutions are LOG RELATIVE RISKS (`exp()` gives the relative risk of failing), the
Weibull `rho` is estimated by Newton alongside the effects, and the frailty variance
comes from a Laplace profile of the marginal likelihood, or is given, as the book does.

```{r}
cows <- data.frame(
  cow  = as.character(8:19),
  herd = as.character(c(1,1,1,1,1,1,2,2,2,2,2,2)),
  ysp  = as.character(c(3,4,1,2,3,1,4,1,2,3,4,2)),
  code = c(0,1,0,1,1,1,1,1,0,1,0,1),
  lpl  = c(40,47,22,28,50,33,49,29,23,37,35,30))
peds <- data.frame(
  animal = as.character(1:19),
  sire = c(rep("0", 7), "1","1","4","4","5","5","1","1","5","5","4","4"),
  dam  = c(rep("0", 7), "2","3","2","9","3","8","6","7","14","6","7","3"))
f12 <- model_survival(lpl ~ herd + ysp + animal(cow), cows, peds, censor = "code",
                      rho = 1, lambda = 1, sigma2 = 0.4, verbose = FALSE)
round(coef(f12, "fixed"), 3)   # Example 16.1: the p.291 solutions of the book
```

## References

The complete list is in the package README. The works this walkthrough leans on directly:
Henderson (1950) for the MME; Henderson (1976) for `A^-1`; Wright (1922) and
Meuwissen and Luo (1992) for inbreeding; Patterson and
Thompson (1971) and Gilmour, Thompson and Cullis (1995) for REML and average information;
Legarra et al. (2015) for metafounders; VanRaden (2008), Aguilar et al. (2010) and
Christensen and Lund (2010) for the single step; Misztal, Legarra and Aguilar (2014) and
Schafer, Katzfuss and Owhadi (2021) for the approximate inverses; Liu et al. (2014), Masuda et al. (2017) and
Vandenplas et al. (2018) for markers as equations; Takahashi, Fagan and Chin (1973) for the
selected inverse; Willham (1972), Kirkpatrick et al. (1990), Schaeffer and Dekkers
(1994), Griffing (1967), Muir and Schinckel (2002), Bijma (2010) and Wade and Quaas (1993) for
the models of section 7; Geman and Geman (1984) and Wang, Rutledge and Gianola (1993, 1994)
for the sampler; Cockerham (1954), Vitezica et al. (2013) and Garcia-Cortes and Toro
(2006) for the declared covariances of section 10; Gianola and Foulley (1983), Foulley,
Gianola and Thompson (1983) and Dempster and Lerner (1950) for the threshold model;
Kachman (1999) for survival. Mrode and Pocrnic (2023) is the textbook behind the
whole sequence, chapter by chapter.
