---
title: "Theory and practice: a genetic evaluation from end to end"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theory and practice: a genetic evaluation from end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2026)
library(BreedingR)
```

This document follows a genetic evaluation in the order it actually happens, and at each
step it says three things: what the matrices are and what they mean, what the algorithm
does with them, and which function runs it. The formulas are the ones the package
computes, not the textbook version they descend from — where the implementation makes a
choice, the choice is stated.

The data for every example is simulated here, so each result has its truth beside it.

```{r}
s <- simulate_breeding(n_founders = 60, n_generations = 3,
                       offspring_per_generation = 150, h2 = 0.4,
                       n_markers = 600, seed = 1)
str(s, max.level = 1)
```

## 1. The model

Every fit in this package is a linear mixed model,

```
y = X b + Z u + e,     u ~ N(0, G),     e ~ N(0, R)
```

where `y` holds the observations, `b` the fixed effects, `u` all the random effects
stacked, and `X` and `Z` their incidence matrices. What separates one model from another
is not the algebra but the shape of `G` and `R`: an animal model puts the pedigree
relationship matrix inside `G`, a repeatability model adds a second block, a multi-trait
model gives `R` a full covariance between traits, a reaction norm gives each animal a
vector of coefficients instead of a scalar.

The package writes all of that in one formula. Fixed terms are bare, `cov()` marks a
fixed covariate, and the marked terms are random: `animal()`, `maternal()`, `sire()`
carry a relationship matrix, `pe()` and `random()` are independent, `rn()` is a random
regression, `indirect()` is the associative effect of pen mates.

**The unit of layout is the covariance group, not the term.** Two terms that share
`group =` go into ONE covariance matrix with the covariance between them estimated:

```
Var(u_g) = C_g (x) K_g
```

where `C_g` is the small covariance among the terms of the group and `K_g` is the
relationship matrix among their levels. This is why direct-maternal, the reaction norm
and the associative model have no dedicated fitter — they are the same engine with a
different incidence and the same penalty.

## 2. The pedigree: A, F, and A inverse

The additive relationship matrix `A` holds twice the kinship between every pair of
animals. It is never built. What the equations need is its inverse, and Henderson (1976)
showed that the inverse has a closed form: for each animal `i` with Mendelian sampling
variance `d_i`,

```
A^-1 = sum_i (1/d_i) v_i v_i',    v_i = +1 at i, -1/2 at sire, -1/2 at dam
```

so `A^-1` is built by adding one small outer product per animal — sparse from the start,
never through a dense `A`. The Mendelian variance depends on how much of the parents'
ancestry is already shared:

```
d_i = 1/2  - 1/4 (F_sire + F_dam)   both parents known
d_i = 3/4  - 1/4 F_known            one parent known
d_i = 1                             neither
```

`F` is the inbreeding coefficient, computed by the Meuwissen and Luo (1992) trace: for
animal `i`, walk the ancestors in decreasing index order accumulating `l_j` contributions,
and `F_i = sum_j l_j^2 d_j - 1`. A max-heap over the ancestors is what makes a single
pass correct — every descendant of an ancestor has a larger index and has therefore
already deposited its share.

```{r}
p <- pedigree(s$pedigree)
head(p, 3)
ai <- a_inverse(s$pedigree)      # i, j, x, n: the lower triangle in triplets
c(animals = ai$n, nonzeros = length(ai$x), mean_F = round(mean(p$F), 4))
```

Two details in the implementation are worth knowing because they are silent when wrong.
A parent that is cited but has no line of its own is an error, not a new founder. And
when sire and dam are the same animal, the coefficients are merged BEFORE the outer
product, which puts `1/d` on that parent's diagonal; emitting the terms one by one into a
half-stored triangle would give `3/4 d^-1` and no error message.

**Metafounders.** When the base population is not one homogeneous unrelated pool —
different lines, different years, animals whose parents are simply unknown — Legarra et
al. (2015) replace the unknown parents with *metafounders*: virtual base animals with a
self-relationship `gamma` that measures how related that base actually was. The
implementation reduces to one line: a metafounder row is preloaded with
`F = gamma - 1` and `d = gamma`, and because every formula above reads only the parents'
`F`, the whole generalization propagates through the Mendelian variances, the trace and
`A^-1` with no special case. `gamma` must be strictly inside `(0, 2)`; that bound is
exactly what keeps `d > 0` for the offspring of two copies of the same metafounder.

```{r}
ped_mf <- s$pedigree
ped_mf$sire[ped_mf$sire == "0"] <- "L1"
ped_mf$dam[ped_mf$dam == "0"] <- "L1"
p_mf <- pedigree(ped_mf, metafounders = "L1", gamma = 0.7)
round(mean(p_mf$F[p_mf$id %in% s$pedigree$id]), 4)   # base relatedness now shows up as F
```

## 3. The mixed model equations

Henderson's equations solve for `b` and `u` at once. Writing `W = [X Z]`, the package
assembles

```
C_s = W'W + blockdiag_g [ kron(s2e * C_g^-1, K_g^-1) ],     rhs = W'y
```

and solves `C_s sol = rhs`. Two conventions matter. The system is scaled by the residual
variance — the penalty carries `s2e * C_g^-1`, not `C_g^-1` — so the solution is the same
BLUP but the matrix is in units of `s2e`. And inside a term, the column of `Z` for
coefficient `a` and level `l` is `a * n_levels + l`: the **level varies fastest**.
Reversing that convention does not break anything visibly — the model still assembles and
converges, and estimates something else.

`W'W` is accumulated one record at a time. Each record touches a handful of columns and
contributes a small clique among them, so the cost is the sum of squared row-lengths, not
anything quadratic in the number of columns. The penalty is added afterwards into the
same triplet list, and the triplet-to-sparse conversion **sums duplicates** — which is
not a convenience but a requirement: in `A^-1` the mate-mate term arrives once per
offspring, and a build where the last write wins would silently discard part of the
matrix.

The restricted likelihood, in the form the MME give:

```
-2logL = (n - p) log s2e + log|G*| + log|C_s| + (y'y - sol'rhs) / s2e
```

The package computes the same quantity a second way, through the dense
`V = Z G Z' + R`, and the tests demand the two agree. They share nothing but the model,
so agreement is evidence; a single path checked against itself would be none.

```{r}
d <- s$data
ev <- eval_internal(y ~ cg + animal(id), d, s$pedigree, theta = c(0.4, 0.6))
c(mme = ev$neg2logl, dense_V = ev$neg2logl_V, difference = abs(ev$neg2logl - ev$neg2logl_V))
```

## 4. Solving: sparse Cholesky and the selected inverse

`C_s` is factored as `L L'` by an up-looking sparse Cholesky, preceded by a minimum
degree ordering that reduces fill-in and by a symbolic analysis that computes the pattern
of `L` once. The symbolic factorization is cached across REML iterations, which is sound
only because the pattern of `C_s` does not depend on the variance components: the
assembly pushes the same slots every time, explicit zeros included.

Prediction error variances need the diagonal of `C_s^-1`, and the full inverse of a
sparse matrix is dense — for a pedigree of any size, unaffordable. Takahashi's identity
gives exactly the elements of the inverse that sit on the pattern of `L`, which is all
the diagonal needs:

```{r}
ch <- sparse_chol(ai)
si <- selected_inverse(ai)
c(logdet = round(ch$logdet, 4), dense_block = ch$bloco)
```

A position outside the pattern is not zero — it is unknown. The function returns only
what it computed.

## 5. Estimating the variance components

REML maximizes the likelihood of error contrasts, which removes the bias that maximum
likelihood carries for the fixed effects (Patterson and Thompson, 1971). The package
maximizes it by average information REML (Gilmour, Thompson and Cullis, 1995), which
needs three ingredients per iteration.

**The score.** For a covariance group with matrix `C_g`, define the quadratic of the
solution and the trace from the selected inverse,

```
Q[a,b] = u_a' K^-1 u_b
T[a,b] = s2e * tr( K^-1 [C_s^-1]_{a,b} )
M_g    = nl * C_g^-1 - C_g^-1 (Q + T) C_g^-1
```

and the derivative of `-2logL` with respect to the element `(i,j)` of `C_g` is `M_g[i,i]`
on the diagonal and `M_g[i,j] + M_g[j,i]` off it, because a symmetric element appears
twice. For the residual,

```
d(-2logL)/ds2e = (n - p - q + trpen)/s2e - e'e/s2e^2
```

where `q` is the number of random columns and `trpen` the trace of the penalty against
`C_s^-1`. The `- q + trpen` is not decoration: the random effects absorb degrees of
freedom too, and a version without it still converges — to the wrong place, reporting
convergence. That is the worst failure mode available, which is why the score is checked
against central finite differences on every parameter.

**The step.** The average information matrix is the average of the observed and expected
information, cheap because it needs no second derivatives, and the step is
`theta <- theta - (AI + lambda diag(AI))^-1 score`. The damping `lambda` starts at 1e-3,
multiplies by ten whenever a candidate is rejected — inadmissible covariance, singular
damped matrix, or a likelihood that got worse — and divides by ten when one is accepted.
Before the AI loop, a few EM rounds run: EM is monotone and stays inside the cone, so it
costs little and lands the search in the right basin.

**Stopping.** Convergence is judged on the relative change of the components,
`sqrt(sum(dtheta^2)/sum(theta^2)) < tol`, never on an absolute threshold on the score,
whose norm grows with the number of records. Standard errors come from the inverse of the
AI matrix, `SE_k = sqrt(2 [AI^-1]_kk)`.

```{r}
fit <- model(y ~ cg + animal(id), d, s$pedigree)
fit
```

Two rules protect the result. A covariance that stops being positive-definite ends the
fit rather than being nudged back, and the starting point always comes from `var(y)`, so
no run inherits a neighbour's answer.

## 6. Genomics: from markers to H

**G.** VanRaden (2008) builds the genomic relationship matrix from centred marker
dosages. With `p_j` the allele frequency at marker `j` and `Z = M - 2p`,

```
G = Z Z' / (2 sum_j p_j (1 - p_j))
```

Missing genotypes are imputed with the marker mean and never with zero — zero is a valid
genotype, and confusing the two changes both the frequencies and every element of `G`
with no error to notice. Markers that are fixed or entirely missing are dropped and
counted.

**A22, and the trap.** The single step needs the pedigree relationships *among the
genotyped animals*, `A22`. That is the inverse of the genotyped block of the inverse, not
the genotyped block of `A^-1` — two matrices of the same shape, both symmetric, only one
correct. The package computes it by a sparse Schur complement,

```
A22^-1 = B22 - B21 B11^-1 B12
```

where `B11` is the non-genotyped block. `B11^-1` is never formed: one sparse
factorization and one triangular solve per genotyped animal do the same work.

**H.** Genomic and pedigree information describe the same animals on two different
scales, so `G` is first brought to the scale of `A22` by an affine adjustment that
matches both the mean diagonal and the mean off-diagonal, then blended:

```
b = (mean_diag(A22) - mean_off(A22)) / (mean_diag(G) - mean_off(G))
a = mean_off(A22) - b * mean_off(G)
G* = (1 - w) (a + b G) + w A22
```

The blend `w` (5 percent by default) keeps `G*` invertible and admits that markers are
not the whole truth. Then the single-step inverse of Aguilar et al. (2010) and
Christensen and Lund (2010) says the whole relationship matrix `H` has an inverse that
differs from `A^-1` only in the genotyped block:

```
H^-1 = A^-1 + [ 0        0            ]
              [ 0   G*^-1 - A22^-1    ]
```

This is the identity that makes the single step practical: nothing dense is needed for
the non-genotyped animals, and the correction is a small block added to a sparse matrix.
It is also self-checking — with `blend = 1`, `G*` becomes `A22` exactly, the correction
block vanishes, and `H^-1` must equal `A^-1` through the whole fit. That collapse is one
of the package's tests.

```{r}
gen <- list(ids = s$genotypes$ids, m = s$genotypes$m)
f_ss <- model(y ~ cg + animal(id), d, s$pedigree, genotypes = gen)
f_id <- model(y ~ cg + animal(id), d, s$pedigree, genotypes = gen, blend = 1)
c(single_step = f_ss$neg2logl, blend_one = f_id$neg2logl, pedigree = fit$neg2logl)
```

**Inverting G\* without paying the cube.** Two approximations are available, and they are
the same idea at different granularity. APY (Misztal, Legarra and Aguilar, 2014) splits
the genotyped into a core and the rest; every non-core animal is regressed on the core,
and what is left is a scalar Mendelian residual:

```
P   = Gcc^-1 Gcn
m_j = g_jj - g_jc P_j
G_APY^-1 = [ Gcc^-1 + P Mnn^-1 P'   -P Mnn^-1 ]
           [ -Mnn^-1 P'              Mnn^-1   ]
```

The Vecchia recursion generalizes it: each animal conditions on its own `k` nearest
relatives among the previous ones rather than on one shared core. Henderson's `A^-1` is
this same recursion with the parents as the conditioning set — exact there, because a
pedigree is Markovian. Schafer, Katzfuss and Owhadi (2021) prove the factor is optimal in
Kullback-Leibler divergence given the pattern, and that enlarging the sets never hurts.

An honest limit: both approximations cut arithmetic, not memory. `G*^-1` is a dense
matrix of the genotyped animals on all three paths.

```{r}
gid <- gen$ids
f_apy <- model(y ~ cg + animal(id), d, s$pedigree, genotypes = gen,
               apy_core = gid[seq(1, length(gid), by = 3)])
f_vec <- model(y ~ cg + animal(id), d, s$pedigree, genotypes = gen, vecchia_k = 60)
round(c(apy = cor(ebv(f_ss, "animal"), ebv(f_apy, "animal")),
        vecchia = cor(ebv(f_ss, "animal"), ebv(f_vec, "animal"))), 4)
```

**Markers as equations.** The other route never forms `G` at all. Liu et al. (2014) write
the equivalent model with the marker effects as unknowns, `u_g = Z g + p_g`, and the
joint precision turns the Schur term into blocks of `A^-1`, so most of the system is the
same sparse matrix the pedigree already gives. `snp_blup()` solves it by preconditioned
conjugate gradients, which need only matrix-vector products, and every application of
`A22^-1` uses the matrix-free identity of Vandenplas et al. (2018):

```
A22^-1 v = A^22 v - A^21 (A^11)^-1 A^12 v
```

one sparse factorization of the non-genotyped block, one triangular solve per iteration.
Nothing dense is ever built. The variance components are given rather than estimated:
this is the solver of routine evaluation, where components come from a periodic REML run.

## 7. What the formula can say

With the covariance group as the unit, several classical models are incidences rather
than fitters.

*Direct-maternal* (Willham, 1972): the animal's own genes and its dam's maternal ability
in one 2x2 matrix, with the direct-maternal covariance — usually negative, and the number
the model exists for.

*Reaction norm* (Kirkpatrick, Lofsvold and Bulmer, 1990): each animal gets a vector of
coefficients on an environmental gradient through `rn(id, base = ...)`, with
`legendre()` building the orthogonal basis. The genetic variance becomes a function of
the environment, which `h2_curve()` reads.

*Indirect genetic effects* (Muir and Schinckel, 2002; Bijma, Muir and Van Arendonk,
2007): the incidence of a record marks the animal's PEN MATES, so a phenotype carries its
own genes plus the associative effects of those it lives with. Direct and indirect share
a group, and their correlation is the parameter that decides whether selecting on own
performance damages the group.

*Multi-trait*: `cbind()` on the left, a full residual covariance between traits, and
missingness handled by the pattern of each record rather than by dropping it.

*AR(1)/CAR(1) residual*: for repeated measures along time, the residual correlation
decays as `rho^|dt|`. The inverse of that correlation matrix is tridiagonal — the
classical Markov result — so the model costs no more than the iid one, and `rho` comes
out with its standard error.

```{r, eval = FALSE}
model(weight ~ cg + animal(id, group = "g") + maternal(dam, group = "g"), d, ped)
model(y ~ cg + rn(id, base = c("phi0", "phi1")) + pe(id), d2, ped)
model(y ~ cg + animal(id, group = "g") + indirect(id, pen = "pen", group = "g"), d, ped)
model_mt(cbind(t1, t2) ~ cg + animal(id), d, ped)
model_ar1(y ~ cg + animal(id) + pe(id), d, ped, subject = "id", time = "day")
```

## 8. Reading the evaluation

The solution of the MME gives the breeding values directly. Their reliability comes from
the prediction error variance, which is where the selected inverse pays off, and the
accuracy of an animal's EBV is

```
acc = sqrt( 1 - PEV / ((1 + F) sigma2_a) )
```

with the inbreeding coefficient in the denominator because an inbred animal has more
additive variance to predict.

```{r}
e <- ebv(fit)
a <- accuracy(fit, s$pedigree)
round(c(cor_with_truth = cor(e[names(s$tbv)], s$tbv),
        mean_accuracy = mean(a), h2 = fit$theta[[1]] / sum(fit$theta)), 3)
```

From there the decisions: `selection_index()` combines traits with economic weights, and
`rank_drift()` answers what an updated evaluation actually changed.

## 9. The Bayesian alternative

The same models can be sampled instead of maximized. `gibbs()` draws all locations in one
block through the same sparse Cholesky, then draws each covariance from its conjugate
inverse-Wishart and the residual from a scaled inverse chi-square, under reference priors.
Where REML gives a point and a standard error from the information matrix, the sampler
gives the whole posterior — worth the cost when a component sits near a boundary, where
the normal approximation behind the standard error is least trustworthy.

```{r}
g <- gibbs(y ~ cg + animal(id), d, s$pedigree, n_iter = 2000, burnin = 400, thin = 4)
data.frame(component = colnames(g$samples),
           posterior_mean = round(unname(g$mean), 3),
           reml = round(unname(fit$theta), 3))
```

## References

The complete list is in the package README. The works this walkthrough leans on directly:
Henderson (1976) for `A^-1`; Meuwissen and Luo (1992) for inbreeding; Patterson and
Thompson (1971) and Gilmour, Thompson and Cullis (1995) for REML and average information;
Legarra et al. (2015) for metafounders; VanRaden (2008), Aguilar et al. (2010) and
Christensen and Lund (2010) for the single step; Misztal, Legarra and Aguilar (2014) and
Schafer, Katzfuss and Owhadi (2021) for the approximate inverses; Liu et al. (2014) and
Vandenplas et al. (2018) for markers as equations; Takahashi, Fagan and Chen (1973) for
the selected inverse; Willham (1972), Kirkpatrick et al. (1990) and Muir and Schinckel
(2002) for the models of section 7.
